---
title: "FinalProject"
author: "Ben Thompson & Cody Coppage"
date: "`r Sys.Date()`"
bibliography: "references.bib"
output: html_document
---

```{r}
# Import Statements

library(dplyr)
#install.packages("fuzzyjoin")
library(fuzzyjoin)
library(lubridate)
library(tidyverse)

```

# Introduction 

Extreme weather events are rapidly emerging as one of the biggest climate-driven ecological stressors worldwide (IPCC 2021). Heat waves, cold snaps, ice storms, and other short-duration extreme weather events are increasing in frequency and severity, and these events can trigger massive ecological impacts in just hours or days (IPCC 2021; Thornton et al. 2023). Oregon is already experiencing rapid swings in temperature and storm intensity, with climate projections suggesting that extreme weather will become increasingly more common in the future (Oregon Climate Change Research Institute 2023). Insects, especially bees, are extremely sensitive to quick shifts in temperature because their flight performance, metabolism, and foraging ability are directly affected by thermal limits (Heinrich 1979; Woods et al. 2019). While a large amount of research shows that pollinator declines are linked to habitat loss and gradual warming trends (Potts et al. 2016), less work has focused on what short-term extreme events do to wild bee populations. Because bees support plant reproduction and agricultural production throughout Oregon, understanding how heat waves, cold snaps, and other extreme weather events affect bee diversity is critical for conservation, ecosystem functioning, and resilience in the face of accelerating climate instability.
Previous research has shown that extreme heat can impair thermoregulation and reduce survival in bee populations (Pereboom & Biesmeijer 2003; Woods et al. 2019), while other work has shown that cold winters can alter pollinator community composition (Franzén & Öckinger 2012). Severe winter storms have been suggested as a risk factor for bees, as cold and ice can reduce floral access and damage habitat (Mishra et al. 2023). These challenges are happening alongside other long-term pressures like habitat loss and floral resource declines (Goulson et al. 2015), which may make bees even more vulnerable when extreme weather occurs. However, not all bee species respond to stress in the same way. One major difference among bees is whether they are specialists or generalists. Some bee genera only use certain host plants, while others can forage from a wide range of flowers (Michener 2000; Wilson & Carril 2015; Danforth et al. 2019). Specialists such as Diadasia, Peponapis, and Perdita are tightly tied to specific pollen sources (Fowler 2016; Praz et al. 2018), so if extreme weather disrupts flowering times or reduces access to those plants, specialists may take the biggest hit. Generalist bees like Bombus, Lasioglossum, and Halictus can use many floral resources (Wilson & Carril 2015; Goulson 2010), which may help them to adapt to quick changes in the environment. Looking at differences between these functional groups may help explain which types are more vulnerable to extreme weather events. 
To address these questions in Oregon, we are using two datasets. First, the Oregon Bee Atlas dataset provides species records and sampling data across Oregon. Second, the NOAA Storm Events Database provides time and location records of extreme heat events, extreme cold events, and ice storms. Because the NOAA data is categorized by county or zone and not by habitat type, our analyses will focus on Oregon-wide patterns rather than trying to isolate habitat differences. By aligning bee sampling dates with storm event windows, we can evaluate whether changes in bee diversity and functional groups occur after extreme weather events such as extreme heat, extreme cold, or ice storm events.
The purpose of our study is to measure how specific extreme weather events influence bee observations across Oregon. We will compare bee data from before and after such weather events, and test whether there are differences in diversity and functional types. This analysis will help fill an information gap, because most research has focused on gradual warming trends or individual mechanisms, rather than short term weather events. The main questions we aim to answer are how does extreme heat, extreme cold, and ice storm events affect bee diversity across Oregon and how do these events affect different functional groups of bees. The answer to these questions will help guide conservation planning under future climate scenarios by identifying which types of extreme weather events are most impactful for bees and how these events affect bee composition. This could help natural resource managers target conservation efforts such as planting strategies, habitat protection, or nest site support in areas most at risk as climate extremes increase.


# Dataset Identification 

The data we are using includes the Oregon Bee Atlas data from 2018-2024 (https://agsci.oregonstate.edu/bee-atlas), and the NOAA Storm Events Database (https://www.google.com/url?sa=t&source=web&rct=j&opi=89978449&url=https://www.ncei.noaa.gov/stormevents/&ved=2ahUKEwjEnvT7iK-RAxUrEzQIHUVmHtkQFnoECAQQAQ&usg=AOvVaw1IQjWMWyvZkgsoVho2eTL5). The storm events database is subset to include only extreme heat events, freezing temperatures, or winter storms in Oregon between 2018 and 2024.

The data skills we are going to use include outer joins, data cleaning with tidyverse, sampling and hypothesis testing.

The metadata (what each column of the data means) for the NOAA Storm Data is as follows:
EVENT_ID: An identification number used by NOAA
CZ_NAME_STR: The location of where the event occurred. Typically a zone or region, such as central Oregon or southern coast. This is the column we used for extracting the county.
BEGIN_DATE: String representing the start date in format DD/MM/YYYY
BEGIN_TIME: Integer data representing time of day, with values from 0 to 2400
EVENT_TYPE: Event type we queried for, includes "Heat", "Extreme cold/wind chill", etc...
MAGNITUDE: Blank column (used for earthquakes, if we had queried for that)
TOR_F_SCALE: Also blank column
DEATHS_DIRECT: Integer data. 0 is used for no deaths
INJURIES_DIRECT: Integer data. 0 is used for no injuries
DAMAGE_PROPERTY_NUM: Integer data. Estimated number of property damage in USD
DAMAGE_CROPS_NUM: Integer data. Estimated amount of crop damage in USD.
STATE_ABBR: String with state abbreviation (all datas are "OR" since that's what we queried for)
CZ_TIMEZONE: All are "PST-8"
MAGNITUDE_TYPE: Blank column, used for other disaster types
EPISODE_ID: Unique ID used by NOAA
CZ_TYPE: Location (C for county, z for forecast zone)
CZ_FIPS: FIPS code for specific location
WFO: Weather Forecast Office in charge of that location
INJURIES_INDIRECT: Amount of indirect injuries recorded
DEATHS_INDIRECT: Amount of indirect deaths recorded
SOURCE: How that data was collected (public, department of highways, etc...)


```{r}
# Importing Oregon Bee Atlas dataset

OBA_dataset <- read.csv("OBA_2018-2024.csv")

#OBA_dataset
```

```{r}
# Importing Storm dataframe from Excel

storm_dataset <- read.csv("storm_data.csv")

#storm_dataset
```

# Data Wrangling/Exploration
```{r}
# Create "GenusSpecies" Column

OBA_dataset <- OBA_dataset %>%
  mutate(GenusSpecies = paste(genus, specificEpithet))

#OBA_dataset
```

```{r}
# Specifying columns to drop from storm events dataframe. These columns are not relevant for our analysis

storm_dataset_cols_to_drop <- c(
  "EVENT_ID", 
  "CZ_NAME_STR", 
  "BEGIN_TIME", 
  "MAGNITUDE", 
  "TOR_F_SCALE", 
  "DEATHS_DIRECT", 
  "INJURIES_DIRECT",
  "DAMAGE_PROPERTY_NUM", 
  "DAMAGE_CROPS_NUM",
  "STATE_ABBR",
  "CZ_TIMEZONE",
  "MAGNITUDE_TYPE",
  "EPISODE_ID",
  "CZ_TYPE",
  "CZ_FIPS",
  "WFO",
  "INJURIES_INDIRECT",
  "DEATHS_INDIRECT",
  "FLOOD_CAUSE",
  "TOR_LENGTH",
  "TOR_WIDTH",
  "BEGIN_RANGE",
  "BEGIN_AZIMUTH",
  "END_RANGE",
  "END_AZIMUTH",
  "END_LOCATION",
  "END_TIME"
  )
```

```{r}
# Dropping the useless columns from storm

storm_dataset <- storm_dataset[, !(names(storm_dataset) %in% storm_dataset_cols_to_drop)]

#storm_dataset
```

```{r}
# Make sure the date columns are date objects 

storm_dataset <- storm_dataset %>%
  mutate(BEGIN_DATE = mdy(BEGIN_DATE))

OBA_dataset <- OBA_dataset %>%
  mutate(verbatimEventDate = mdy(verbatimEventDate))

#OBA_dataset
```

```{r}
# Join Tables

joined_data <- fuzzy_left_join(OBA_dataset, storm_dataset,
  by = c("county" = "COUNTY",
         "verbatimEventDate" = "BEGIN_DATE"), 
  match_fun = list(`==`, function(bee_date, storm_date) {
    bee_date >= storm_date & bee_date <= storm_date + 30}))

#joined_data
```

```{r}
# Count of each unique storm that occurred before a bee capture instance

joined_data %>%
  count(EVENT_TYPE)
#joined_data
```


```{r}
# Add Control column. True if control data, false if a storm instance
joined_data <- joined_data %>%
  mutate(Control = is.na(EVENT_TYPE))
#joined_data
```


```{r}
# Remove rows with no species data

joined_data <- joined_data %>%
  filter(specificEpithet != "")
joined_data <- joined_data %>%
  filter(specificEpithet != " ")
```
## Data Analysis & Hypothesis Testing - AB Test for Storm vs No Storm

Before we do our AB test to determine whether extreme weather impacts bee diversity, we need to come up with a null hypothesis to either reject or not. Or null hypothesis will be: Bees are equally diverse, according to the shannon diversity index, in areas which have recently been impacted by either extreme heat or freezing temperatures. If we get a p value of less than 0.05, we will reject this null and assume that the weather does impact bees. 

To do our analysis, we will create a function to compute Shannon Diversity, then find our observed statistic for shannon diversity of bees collected after certain extreme weather events. Our actual observed statistic will be the difference in Shannon Diversites between bees collected outside of the storm timeframe (30 days) and bees collected within 30 days of a storm. After that, we will collect some arbitrary number (n_iter = 1000) to iterate through. In each iteration, we will shuffle the control column and re-compute the differences. Then we will see if our observed statistic falls within ght 95% range of the sampled statistics. 

```{r}
# Find Species Richness
S <- length(unique(joined_data$specificEpithet))
S
```


```{r}
# Natural Log of Species Richness (Necessary for Species evenness, but not Shannon Diversity)
ln_s <- log(S)
ln_s
```


```{r}
# Calculate Proportion of Each Species

species_counts <- joined_data %>%
  count(specificEpithet, name = 'n')

proportions <- c()

j <- 1

for (i in species_counts$n) {
  proportions[j] = (i / sum(species_counts$n))
  j <- j+1
}
#proportions

```


```{r}
# Calculate Overall Shannon Diversity Index
# H = – Summation[P(i) * lnP(i)].

summation <- 0
for (p in proportions) {
  summation <- summation + (p * log(p))
}

shannon_diversity_index <- (summation * -1)

```

We found out there is already a package to compute shannon diversity, but we will still create a reproducible function to compute it as well so that we understand better how it runs. However, we will still test it here to make sure our numbers line up.

```{r}
# Calculate Shannon Diversity Index Another Way

library(vegan)

shannon_diversity_index_2 <- joined_data %>%
  summarize(diversity = diversity(table(specificEpithet)))
shannon_diversity_index_2

```


```{r}
# Calculate Shannon Diversity for Control Group

joined_data
control_table <- joined_data %>%
  filter(Control == TRUE)

S_control <- length(unique(control_table$specificEpithet))

species_counts_control <- control_table %>%
  count(specificEpithet, name = 'n')

proportions_control <- c()

j <- 1

for (i in species_counts_control$n) {
  proportions_control[j] = (i / sum(species_counts_control$n))
  j <- j+1
}
proportions_control

summation_control <- 0
for (p in proportions_control) {
  summation_control <- summation_control + (p * log(p))
}

shannon_diversity_index_control <- (summation_control * -1)

shannon_diversity_index_control
```




```{r}
# Calculate Shannon Diversity of Treatment Group
joined_data
not_control_table <- joined_data %>%
  filter(Control == FALSE)
control_table <- joined_data %>%
  filter(Control == TRUE)

S_not_control <- length(unique(not_control_table$specificEpithet))

species_counts_not_control <- not_control_table %>%
  count(specificEpithet, name = 'n')

proportions_not_control <- c()

j <- 1

for (i in species_counts_not_control$n) {
  proportions_not_control[j] = (i / sum(species_counts_not_control$n))
  j <- j+1
}
proportions_not_control

summation_not_control <- 0
for (p in proportions_not_control) {
  summation_not_control <- summation_not_control + (p * log(p))
}

shannon_diversity_index_treatment <- (summation_not_control * -1)

shannon_diversity_index_treatment


```
```{r}
# Create function to compute Shannon Diversity (reproducible)

shannon_diversity <- function(df, col) {
  col <- deparse(substitute(col))

  S <- length(unique(df[[col]]))
  
  species_counts <- df %>%
    count(.data[[col]], name = 'n')
  
  proportions <- c()
  
  j <- 1

  for (i in species_counts$n) {
    proportions[j] = (i / sum(species_counts$n))
    j <- j+1
  }
  
  summation <- 0
  for (p in proportions) {
    summation <- summation + (p * log(p))
  }

  H <- (summation * -1)
  
  return(H)
  
}

shannon_diversity(not_control_table, specificEpithet)

```
So, the Shannon diversity for "storm bees" was 4.3852, and the shannon diversity for bees captured in normal circumstances was 4.646. The diversity is lower for bees captured around storms (Extreme heat or freezing temperatures), but is it significantly so? We will use A/B testing to determine if so.


```{r}
# Set seed for reproducibility, and the number of iterations to shuffle
set.seed(1)   
n_iterations <- 1000

```




```{r}
# Perform AB Testing

shuffle_results <- c()
joined_data_filtered_true <- joined_data %>%
  filter(Control == TRUE)
joined_data_filtered_false <- joined_data %>%
  filter(Control == FALSE)


for (i in 1:n_iterations) {
  joined_data_sampled <- joined_data %>%
    mutate(shuffled = sample(Control))
  temp <- joined_data_sampled %>%
    filter(shuffled == TRUE)
  temp2 <- joined_data_sampled %>%
    filter(shuffled == FALSE)
  shuffle_results[i] <- shannon_diversity(temp, specificEpithet) - shannon_diversity(temp2, specificEpithet)
}


sort(shuffle_results)


```

```{r}
# Assign P Value (Probability that our observed shannon diversity difference between storm/no storm for bees is at least that small if Null were true)

p_value <- mean(shuffle_results <= shannon_diversity(not_control_table, specificEpithet) - shannon_diversity(control_table, specificEpithet))
p_value
```
A P value of 0 means that we reject the null hypothesis. This means that the extreme weather events do impact bee diversity, making the bees less diverse in areas that recently experienced either extreme heat or freezing temperatures. 



```{r}
# Histogram of Shuffled Results
ggplot(data.frame(x = shuffle_results), aes(x)) +
  geom_histogram() +
  ggtitle("Distribution of Shuffled AB Test Shannon Diversities") + 
  xlab("Shuffled Shannon Diversity Values") +
  ylab("Frequency")

shannon_diversity(joined_data_filtered_true, specificEpithet) - shannon_diversity(joined_data_filtered_false, specificEpithet)
```

```{r}
# Histogram of Shuffled Results, plus vertical line showing non-shuffled diversity
ggplot(data.frame(x = shuffle_results), aes(x)) +
  geom_histogram(fill = "#1E90FF", color = "black", alpha = 0.8, binwidth = 0.009) +
  ggtitle("Distribution of Shuffled AB Test Shannon Diversities + Line for Observed Statistic") + 
  geom_vline(xintercept = -(shannon_diversity(joined_data_filtered_true, specificEpithet) - shannon_diversity(joined_data_filtered_false, specificEpithet)), color = "red", linetype = "dashed", size = 1) + 
  xlab("Shuffled Shannon Diversity Values") +
  ylab("Frequency") + 
  theme(plot.title = element_text(size = 10, face = "bold", hjust = 0.5))

```
As we can see, our observed statistic (Shannon diversity of storm bees) falls well outside of the limits of our shuffled results, meaning it is statistically significant.

## Data Analysis: Does freezing temperatures impact bee diversity?

Now, we want to do another AB test, but this time just to see if frost/freeze impacts bee diversity (ignoring extreme heat). Our null hypothesis this time will be: Freezing temperatures do not impact bee diversity according to the Shannon diversity index. 

```{r}
# Create new table to analyze freezing temperatures

joined_data_frost <- joined_data %>%
  mutate(frost = coalesce(EVENT_TYPE == "Frost/Freeze", FALSE))
joined_data_frost
```

```{r}
# Calculate Observed Statistic
joined_data_frost_filtered_true <- joined_data_frost %>%
  filter(frost == TRUE)
joined_data_frost_filtered_false <- joined_data_frost %>%
  filter(frost == FALSE)
shannon_diversity(joined_data_frost_filtered_true, specificEpithet) - shannon_diversity(joined_data_frost_filtered_false, specificEpithet)
```


```{r}
# Perform AB Test
shuffled_frost_results <- c()
for (i in 1:n_iterations) {
  joined_data_frost <- joined_data_frost %>%
    mutate(frost_shuffled = sample(frost))
  temp <- joined_data_frost %>%
    filter(frost_shuffled == TRUE)
  temp2 <- joined_data_frost %>%
    filter(frost_shuffled == FALSE)
  shuffled_frost_results[i] <- shannon_diversity(temp, specificEpithet) - shannon_diversity(temp2, specificEpithet)
}

sort(shuffled_frost_results)
```


```{r}
# Compute P Value for freezing temperatures
p_value_frost <- mean(shuffled_frost_results <= shannon_diversity(joined_data_frost_filtered_true, specificEpithet) - shannon_diversity(joined_data_frost_filtered_false, specificEpithet))
p_value_frost
```

```{r}
# Histogram for frost results
ggplot(data.frame(x = shuffled_frost_results), aes(x)) +
  geom_histogram(fill = "skyblue", color = "black", alpha = 0.8, binwidth = 0.009) +
  ggtitle("Distribution of Shuffled Frost Results + Line for Observed Statistic") + 
  geom_vline(xintercept = shannon_diversity(joined_data_frost_filtered_true, specificEpithet) - shannon_diversity(joined_data_frost_filtered_false, specificEpithet), color = "red", linetype = "dashed", size = 1) + 
  xlab("Shuffled Shannon Diversity Values") +
  ylab("Frequency") +
  theme(plot.title = element_text(size = 13, face = "bold", hjust = 0.5))


```
As we see here, the p value is once again 0. This means that we reject the null, and that freezing temperatures does negatively impace the shannon diversity of bees.



## Data Analysis - AB Test for Extreme Heat
For our last AB test, we will look at whether or not extreme heat impacts bee diversity. Our null hypothesis: Extreme heat events do not impact the diversity of bees. 
```{r}
# Create new table to analyze extreme heat

joined_data_heat <- joined_data %>%
  mutate(heat = coalesce(EVENT_TYPE == "Excessive Heat", FALSE))
joined_data_heat
```

```{r}
# Calculate Observed Statistic
joined_data_heat_filtered_true <- joined_data_heat %>%
  filter(heat == TRUE)
joined_data_heat_filtered_false <- joined_data_heat %>%
  filter(heat == FALSE)
shannon_diversity(joined_data_heat_filtered_true, specificEpithet) - shannon_diversity(joined_data_heat_filtered_false, specificEpithet)

```

```{r}
# Perform AB Test
shuffled_heat_results <- c()
for (i in 1:n_iterations) {
  joined_data_heat_temp <- joined_data_heat %>%
    mutate(heat_shuffled = sample(heat))
  temp <- joined_data_heat_temp %>%
    filter(heat_shuffled == TRUE)
  temp2 <- joined_data_heat_temp %>%
    filter(heat_shuffled == FALSE)
  shuffled_heat_results[i] <- shannon_diversity(temp, specificEpithet) - shannon_diversity(temp2, specificEpithet)
}

sort(shuffled_heat_results)

```

```{r}
# Compute P Value for Shannon Diversity of Heat Bees
p_value_heat <- mean(shuffled_heat_results <= shannon_diversity(joined_data_heat_filtered_true, specificEpithet) - shannon_diversity(joined_data_heat_filtered_false, specificEpithet))
p_value_heat

```
```{r}
# Histogram for Sampled Heat Shannon Diversity Differences
ggplot(data.frame(x = shuffled_heat_results), aes(x)) +
  geom_histogram(fill = "orange", color = "black", alpha = 0.8, binwidth = 0.009) +
  ggtitle("Distribution of Shuffled Heat Results + Line for Observed Statistic") + 
  geom_vline(xintercept = shannon_diversity(joined_data_heat_filtered_true, specificEpithet) - shannon_diversity(joined_data_heat_filtered_false, specificEpithet), color = "red", linetype = "dashed", size = 1) + 
  xlab("Shuffled Shannon Diversity Values") +
  ylab("Frequency") + 
  theme(plot.title = element_text(size = 13, face = "bold", hjust = 0.5))

```
Once again, we reject the Null hypothesis. We also see based on the difference in shannon diversities, that heat actually has more impact on the diversity than the extreme cold!

```{r}
diversities <- data.frame(
  Category = c("Storm"),
  Frost = shannon_diversity(joined_data_frost_filtered, specificEpithet),
  Heat = shannon_diversity(joined_data_heat_filtered, specificEpithet),
  "No Storm" = shannon_diversity_index_control
)

diversities_long <- diversities %>%
  pivot_longer(cols = -Category, names_to = "Type", values_to = "Value")

ggplot(diversities_long, aes(x = Type, y = Value, fill = Type)) +
  geom_bar(stat = "identity", show.legend = FALSE, color = "black", alpha = 0.8) +
  ggtitle("Shannon Diversity by Storm") + 
  xlab("Storm Type") +
  ylab("Shannon Diversity") + 
  theme(plot.title = element_text(size = 13, face = "bold", hjust = 0.5)) +
  scale_fill_manual(values = c("Frost" = "blue", "Heat" = "orange", "No Storm" = "grey")) 



```

# Visualizations

The following figures show how diversity and community structure are affected by the sampled extreme weather events. 

## Effect on Shannon diversity

To visualize how diversity is affected by the sampled weather events we first had to find the dates of our sampled weather events. Once dates of events were found we used a for loop to subset the data to samples within 30 day windows before and after events and then apply a Shannon diversity function to the subsetted data to give us a data frame that listed Shannon diversity values before and after events. For ease of visualization, we took the mean of these diversity values per event type and plotted them on a bar chart. We chose a bar plot because it's the clearest way to compare diversity values before and after events by event type. Visualization shows that there is a noticeable difference in Shannon diversity before and after events with lower values following extreme heat and cold events. This indicates that extreme weather events lead to a decrease in bee diversity across Oregon following these events.

```{r}
# Find dates of events
Event_Date_Data <- joined_data %>%
  filter(!is.na(EVENT_TYPE)) %>%
  select(BEGIN_DATE, EVENT_TYPE) %>%
  distinct()
```

```{r}
# Create subsets of bees 30-days pre/post events and compute Shannon diversity for each
n_events <- nrow(Event_Date_Data)

pre_div  <- numeric(n_events)
post_div <- numeric(n_events)

for (i in 1:n_events) {
  event_day <- Event_Date_Data$BEGIN_DATE[i]
  
  # Bees 30 days before
  pre_bees <- joined_data %>%
    filter(verbatimEventDate >= event_day - 30,
           verbatimEventDate <  event_day)
  
  # Bees 30 days after
  post_bees <- joined_data %>%
    filter(verbatimEventDate >  event_day,
           verbatimEventDate <= event_day + 30)
  
  # Counts per species
  pre_tab  <- table(pre_bees$GenusSpecies)
  post_tab <- table(post_bees$GenusSpecies)
  
  # Calculate Shannon diversity
  pre_div[i]  <- diversity(pre_tab)
  post_div[i] <- diversity(post_tab)
}
```


```{r}
# Comparison Table
diversity_results <- data.frame(EVENT_DATE = Event_Date_Data$BEGIN_DATE,
                                EVENT_TYPE = Event_Date_Data$EVENT_TYPE,
                                shannon_pre30 = pre_div,
                                shannon_post30 = post_div,
                                diff_post_minus_pre = post_div - pre_div)

diversity_results
```

```{r}
# Find the means for each event type
event_type_summary <- diversity_results %>%
  group_by(EVENT_TYPE) %>%
  summarise(mean_pre  = mean(shannon_pre30),
            mean_post = mean(shannon_post30),
            mean_difference = mean(diff_post_minus_pre),
            n_events = n())

event_type_summary
```
```{r}
# Bar graph showing differences in pre/post event, grouped by event type
# Reshape data for plotting
plot_data <- event_type_summary %>%
  select(EVENT_TYPE, mean_pre, mean_post) %>%
  pivot_longer(cols = c(mean_pre, mean_post), names_to = "period", values_to = "diversity")
# Change order of data for plot
plot_data$period <- factor(plot_data$period,
                           levels = c("mean_pre", "mean_post"),
                           labels = c("Pre-Event", "Post-Event"))
# Plot data
ggplot(plot_data, aes(x = EVENT_TYPE, y = diversity, fill = period)) +
  geom_col(position = "dodge",color = "black", alpha = 0.8) +
  ggtitle("Bee Diversity Before vs. After Storm Events") + 
  xlab("Event Type") + 
  ylab("Mean Shannon Diversity") + 
  theme(plot.title = element_text(size = 13, face = "bold", hjust = 0.5)) + 
  #theme_minimal() +
  labs(fill = "Sampling Period") +
  scale_fill_manual(values=c("magenta", "green"))

```

## Effect on community structure

To compare community structure before and after extreme weather events bees were classified as specialists or generalists using genus-level traits derived from established ecological references (Michener 2000; Wilson & Carril 2015; Danforth et al. 2019). Genera with strong evidence of specialization (e.g., Diadasia, Peponapis, Perdita, Chelostoma, Dufourea) were categorized as specialists based on associations documented by Fowler (2016) and Praz et al. (2018) and genera (e.g., Bombus, Lasioglossum, Halictus) were categorized as generalists through documentations by Wilson & Carril (2015) and Goulson (2010). Bee identities were additionally cross-referenced with the USGS Native Bee Inventory dataset (Fowler & Droege 2020), which is a widely used North American bee monitoring program. Genera were only assigned to specialist or generalist categories when consistent agreement existed in the literature. Ambiguous genera were excluded from assignment to avoid confusion or bias in the data. We decided to use genera instead of species-specific traits due to time constraints, complexity, and dissagreement in the literature. Once genera were categorized we created a new data table and used a for loop to subset the data for bee samples within 30 day windows before and after events, count the number of samples for each genera, and then find the overall proportions relative to specialists for ease of plotting. Once this was completed we summarized the data by event type and plotted the data in a bar chart. The bar chart shows that there is a  shift in community structure towards generalist genera which indicates that extreme weather events cause a noticeable difference in community structure with specialist genera not able to survive as well as generalist species after these events. 

```{r}
# Create object vectors for Genera dependent on if they are specialists or generalists. Genera that have species that are both are excluded for clarity

specialist_genera <- c("Diadasia", "Peponapis", "Hesperapis", "Dufourea", "Chelostoma", "Perdita", "Colletes", "Melissodes")

generalist_genera <- c("Bombus","Halictus","Apis","Lasioglossum","Eucera","Ceratina","Agapostemon",
                       "Xylocopa","Osmia","Anthophora","Megachile","Hylaeus","Calliopsis","Heriades",
                       "Ashmeadiella","Atoposmia","Trachusa","Dianthidium","Tetraloniella",
                       "Pseudoanthidium","Protosmia","Anthidiellum","Hoplitis","Habropoda")
# For clarity
excluded_genera <- c("Andrena","Micralictoides","Stelis","Sphecodes","Melecta","Coelioxys","Epeolus",
                     "Neolarra","Oreopasites","Nomia","Triepeolus","Biastes","Epimelissodes",
                     "Protohalonia","Dioxys","Zacosmia","Holcopasites","Brachymelecta")

```

```{r}
# Make a new data frame and loop to analyze community structure
# Create data table
structure_table <- joined_data %>%
  filter(!is.na(EVENT_TYPE)) %>%
  select(BEGIN_DATE, EVENT_TYPE) %>%
  distinct()
n_events_s <- nrow(structure_table)

```

```{r}
# Create Vectors to store info from for loop
pre_spec        <- numeric(n_events)
pre_gen         <- numeric(n_events)
post_spec       <- numeric(n_events)
post_gen        <- numeric(n_events)
pre_prop_spec   <- numeric(n_events)
post_prop_spec  <- numeric(n_events)
```

```{r}
# For loop to find proportions of generalist vs specialist bees and how extreme weather affected them

for (i in 1:n_events_s) {
  
  event_day <- structure_table$BEGIN_DATE[i]
  
  # Bees 30 days before event 
  pre_bees <- joined_data %>%
    filter(verbatimEventDate >= event_day - 30,
           verbatimEventDate <  event_day) %>%
    filter(genus %in% c(specialist_genera, generalist_genera))
  
  # Bees 30 days after event
  post_bees <- joined_data %>%
    filter(verbatimEventDate >  event_day,
           verbatimEventDate <= event_day + 30) %>%
    filter(genus %in% c(specialist_genera, generalist_genera))
  
  pre_spec[i]  <- sum(pre_bees$genus %in% specialist_genera)
  pre_gen[i]   <- sum(pre_bees$genus %in% generalist_genera)
  post_spec[i] <- sum(post_bees$genus %in% specialist_genera)
  post_gen[i]  <- sum(post_bees$genus %in% generalist_genera)

  pre_total  <- pre_spec[i]  + pre_gen[i]
  post_total <- post_spec[i] + post_gen[i]
  
  # Found proportions kept giving 0 values so need to make sure that we aren't dividing by 0
  if (pre_total > 0) {
    pre_prop_spec[i] <- pre_spec[i] / pre_total
  } else {
    pre_prop_spec[i] <- NA
  }
  if (post_total > 0) {
    post_prop_spec[i] <- post_spec[i] / post_total
  } else {
    post_prop_spec[i] <- NA
  }
}
```


```{r}
# Comparison table
structure_results <- data.frame(EVENT_DATE = structure_table$BEGIN_DATE, 
                                EVENT_TYPE = structure_table$EVENT_TYPE,
                                pre_spec = pre_spec,
                                pre_gen = pre_gen,
                                post_spec = post_spec,
                                post_gen = post_gen,
                                pre_prop_spec = pre_prop_spec,
                                post_prop_spec = post_prop_spec,
                                diff_prop_spec = post_prop_spec - pre_prop_spec)

structure_results
```

```{r}
# Plot the data
# Summarize by event type
structure_summary <- structure_results %>%
  group_by(EVENT_TYPE) %>%
  summarise(mean_pre_prop_spec  = mean(pre_prop_spec,  na.rm = TRUE),
    mean_post_prop_spec = mean(post_prop_spec, na.rm = TRUE))

structure_plot_data <- structure_summary %>%
  pivot_longer(cols = c(mean_pre_prop_spec, mean_post_prop_spec), names_to = "period", values_to = "prop_spec")

structure_plot_data$period <- factor(structure_plot_data$period, levels = c("mean_pre_prop_spec", "mean_post_prop_spec"), labels = c("Pre-Event", "Post-Event"))

structure_plot_data_filtered <- structure_plot_data %>%
  filter(EVENT_TYPE != "Winter Storm")

ggplot(structure_plot_data_filtered, aes(x = EVENT_TYPE, y = prop_spec, fill = period)) +
  geom_col(position = "dodge", color = "black", alpha = 0.8) +
  theme_minimal() +
  ggtitle("Bee Community Structure (Specialists vs Generalists) Before and After Extreme Weather Events") + 
  xlab("Event Type") + 
  ylab("Mean Proportion Specialists") + 
  labs(fill = "Sampling Period") +
  theme(plot.title = element_text(size = 10, face = "bold", hjust = 0.5)) + 
  scale_fill_manual(values=c("magenta", "green"))
```

```{r}
# Calculate Percentage Decrease in Specialist Proportion After Heat
((structure_plot_data_filtered$prop_spec[1] - structure_plot_data_filtered$prop_spec[2]) / structure_plot_data_filtered$prop_spec[1]) * 100
```
```{r}
# Calculate Percentage Decrease in Specialist Proportion After Freezing Temperatures
((structure_plot_data_filtered$prop_spec[3] - structure_plot_data_filtered$prop_spec[4]) / structure_plot_data_filtered$prop_spec[3]) * 100
```
# Limitations / Potential Problems
Our storm data only had around 600 instances, and some counties were not represented at all, such as Baker, Umatilla, and Coos. So when we analyze diversity among bees after storms, the lack of diversity in location may actually impact our diversity in bee community. Additionally, our storm data is not all-encompassing. There were only 600 instances of combined freezing temperatures and heat (over 100 degrees F). This is obviously not all instances of extreme temperatures recorded, so our data is not taking into account all instances of extreme temperature. The discrepency found between counties with bee records and storms is obvious if you look at the ArcGIS Pro layout found in the repo.

After speaking with Dr. Lincoln Bartlett, one of the researchers in charge of maintaining the Oregon Bee Atlas data, we learned that the Shannon Diversity index is not typically very strong at calculating diversities over such a large dataset. With so many bees in the data, the diversity is not great because the values will almost always be very large (between 4 and 5) and therefore it will be harder to find diversities that small in subsets of it like the storms. This is likely why our P values were all 0.

Because the NOAA data does not contain specific locations (xy coordinates) we were not able to compute where and when exactly temperature events occurred. For example, county regions are very large, and climate may be much different at the northern tip of a country than the southern tip. Perhaps certain bees we recorded as "storm bees" were not actually storm bees because they were in a separate part of the county. 

With bees collected by hand in the field, it is natural that extreme weather events would impact the number of people going out into the field to collect data. Perhaps bees do not actually get impacted by small term weather events, but the collection is impacted because nobody wants to go out in the field in the middle of a hot summer or a freezing winter. 


# Conclusion


Our analysis reveals evidence that extreme temperatures in either direction do lead to bees being both less diverse, and to become more specialized. We concluded that, with P=0 confidence, that bee communities are less diverse according to the Shannon diversity index after extreme temperature events. However, because of dataset limitations and external factors surrounding the hand-collected data, we can not be certain of any conclusions. More analysis, with higher resolution temperature data, would be needed to draw any conclusions.


# References (Temporary spot. Needs to be changed to BibTex)

Danforth, B. N., Minckley, R. L., Neff, J. L., & Fawcett, F. (2019). The solitary bees: Biology, evolution, conservation. Princeton University Press.

Fowler, J. (2016). Specialist Bees of the Northeast: Host Plants and Habitat Conservation. Northeastern Naturalist, 23(2), 305–320. https://doi.org/10.1656/045.023.0210

Franzén, M., & Öckinger, E. (2012). Climate-driven changes in pollinator communities: The return of the dark side of cold winters. Journal of Insect Conservation, 16(2), 215–222. https://doi.org/10.1007/s10841-011-9410-y

Goulson, D. (2010). Bumblebees: Behaviour, ecology, and conservation. Oxford University Press.

Goulson, D., Nicholls, E., Botías, C., & Rotheray, E. L. (2015). Bee declines driven by combined stress from parasites, pesticides, and lack of flowers. Science, 347(6229), 1255957. https://doi.org/10.1126/science.1255957

Heinrich, B. (1979). Bumblebee economics. Harvard University Press.

Intergovernmental Panel on Climate Change. (IPCC) (2021). Climate change 2021: The physical science basis (V. Masson-Delmotte, P. Zhai, A. Pirani, et al., Eds.). Cambridge University Press. https://doi.org/10.1017/9781009157896

Michener, C. D. (2000). The bees of the world. Johns Hopkins University Press.

Mishra, M., Bhunia, P., Shubham, S., Sen, R., Bhunia, R., Gulia, J., Mondal, T., Zidane, Z., Saha, T., Chacko, A., Raj, R., Vyshak, S. S., & Kaushal, L. (2023). The impact of weather change on honey bee populations and disease. Journal of Advanced Zoology, 44(S7), 180–190. https://doi.org/10.17762/jaz.v44iS7.2755

Oregon Climate Change Research Institute. (2023). The Fifth Oregon Climate Assessment. Oregon State University, Corvallis, OR. https://www.climatechange.oregonstate.edu/oregon-climate-assessments

Pereboom, J. J., & Biesmeijer, J. C. (2003). Thermal constraints for stingless bee foragers: The importance of body size and coloration. Oecologia, 137(1), 42–50. https://doi.org/10.1007/s00442-003-1333-6

Potts, S. G., Imperatriz-Fonseca, V., Nguyen, H. T., Bartuska, A., Medellín, R. A., Baste, I. A., Oteng-Yeboah, A., Watson, R., et al. (Eds.). (2016). The assessment report of the Intergovernmental Science-Policy Platform on Biodiversity and Ecosystem Services on pollinators, pollination and food production. Secretariat of the Intergovernmental Science-Policy Platform on Biodiversity and Ecosystem Services.

Praz, C. J., Müller, A., & Dorn, S. (2018). Host-plant specialization in solitary bees: A case for polylecty evolution. Biological Reviews, 93(4), 1693–1716. https://doi.org/10.1111/brv.12414

Russo, L. (2016). Positive and negative impacts of non-native bees on native ecosystems. Annual Review of Entomology, 61, 13–31. https://doi.org/10.1146/annurev-ento-010715-023719

Thornton, P. K., Whitfield, S., & Challinor, A. J. (2023). Climate extremes and their implications for global food systems. Annual Review of Environment and Resources, 48, 203–229. https://doi.org/10.1146/annurev-environ-120120-021013

Wilson, J. S., & Carril, O. M. (2015). The bees in your backyard: A guide to North America’s bees. Princeton University Press.

Woods, H. A., Dillon, M. E., & Pincebourde, S. (2019). The roles of microclimatic diversity and of behavior in mediating the responses of ectotherms to climate change. Journal of Thermal Biology, 84, 80–92. https://doi.org/10.1016/j.jtherbio.2019.06.003
