---
title: "FinalProject"
author: "Ben Thompson & Cody Coppage"
date: "`r Sys.Date()`"
bibliography: "references.bib"
output: html_document
---

```{r}
# Import Statements

library(dplyr)
#install.packages("fuzzyjoin")
library(fuzzyjoin)
library(lubridate)
library(tidyverse)




```

# Introduction 

Extreme weather events are rapidly emerging as one of the biggest climate-driven ecological stressors worldwide (IPCC 2021). Heat waves, cold snaps, ice storms, and other short-duration extreme weather events are increasing in frequency and severity, and these events can trigger massive ecological impacts in just hours or days (Bilton et al. 2019; Thornton et al. 2023). Oregon is already experiencing rapid swings in temperature and storm intensity, with climate projections suggesting that extreme weather will become increasingly more common in the future (Oregon Climate Change Research Institute 2023). Insects, especially bees, are extremely sensitive to quick shifts in temperature because their flight performance, metabolism, and foraging ability are directly affected by thermal limits (Heinrich 1979; Woods et al. 2019). While a large amount of research shows that pollinator declines are linked to habitat loss and gradual warming trends (Potts et al. 2016), less work has focused on what short-term extreme events do to wild bee populations. Because bees support plant reproduction and agricultural production throughout Oregon, understanding how heat waves, cold snaps, and other extreme weather events affect bee diversity is critical for conservation, ecosystem functioning, and resilience in the face of accelerating climate instability.
Previous research has shown that extreme heat can cause bees to overheat and die by shutting down normal thermoregulation (Pereboom & Biesmeijer 2003; Woods et al. 2019). Other work has shown that sudden freezing temperatures can kill overwintering bees or prevent them from emerging from nests (Franzén & Öckinger 2012). Severe winter storms have also been linked to declines in bee populations because ice can block access to flowers or physically damage habitat (Riley et al. 2018). These challenges are happening alongside other long-term pressures like habitat loss and floral resource declines (Goulson et al. 2015), which may make bees even more vulnerable when extreme weather occurs. However, not all bee species respond to stress in the same way. One major difference among bees is whether they are specialists or generalists. Some bee genera only use certain host plants, while others can forage from a wide range of flowers (Michener 2000; Wilson & Carril 2015; Danforth et al. 2019). Specialists such as Diadasia, Peponapis, and Perdita are tightly tied to specific pollen sources (Fowler & Droege 2020; Praz et al. 2018), so if extreme weather disrupts flowering times or reduces access to those plants, specialists may take the biggest hit. Generalist bees like Bombus, Lasioglossum, and Halictus can use many floral resources (Wilson & Carril 2015; Goulson 2010), which may help them to adapt to quick changes in the environment. Looking at differences between these functional groups may help explain which types are more vulnerable to extreme weather events. 
To address these questions in Oregon, we are using two datasets. First, the Oregon Bee Atlas dataset provides species records and sampling data across Oregon. Second, the NOAA Storm Events Database provides time and location records of extreme heat events, extreme cold events, and ice storms. Because the NOAA data is categorized by county or zone and not by habitat type, our analyses will focus on Oregon-wide patterns rather than trying to isolate habitat differences. By aligning bee sampling dates with storm event windows, we can evaluate whether changes in bee diversity and functional groups occur after extreme weather events such as extreme heat, extreme cold, or ice storm events.
The purpose of our study is to measure how specific extreme weather events influence bee observations across Oregon. We will compare bee data from before and after such weather events, and test whether there are differences in diversity and functional types. This analysis will help fill an information gap, because most research has focused on gradual warming trends or individual mechanisms, rather than short term weather events. The main questions we aim to answer are how does extreme heat, extreme cold, and ice storm events affect bee diversity across Oregon and how do these events affect different functional groups of bees. The answer to these questions will help guide conservation planning under future climate scenarios by identifying which types of extreme weather events are most impactful for bees and how these events affect bee composition. This could help natural resource managers target conservation efforts such as planting strategies, habitat protection, or nest site support in areas most at risk as climate extremes increase.


# Dataset Identification 

```{r}
# Importing Oregon Bee Atlas dataset

OBA_dataset <- read.csv("OBA_2018-2024.csv")

OBA_dataset
```

```{r}
# Importing Storm dataframe from Excel

storm_dataset <- read.csv("storm_data.csv")

storm_dataset
```

# Data Wrangling/Exploration
```{r}
# Create "GenusSpecies" Column
OBA_dataset <- OBA_dataset %>%
  mutate(GenusSpecies = paste(genus, specificEpithet))
OBA_dataset
```

```{r}
# Specifying columns to drop from storm events dataframe. These columns are not relevant for our analysis

storm_dataset_cols_to_drop <- c(
  "EVENT_ID", 
  "CZ_NAME_STR", 
  "BEGIN_TIME", 
  "MAGNITUDE", 
  "TOR_F_SCALE", 
  "DEATHS_DIRECT", 
  "INJURIES_DIRECT",
  "DAMAGE_PROPERTY_NUM", 
  "DAMAGE_CROPS_NUM",
  "STATE_ABBR",
  "CZ_TIMEZONE",
  "MAGNITUDE_TYPE",
  "EPISODE_ID",
  "CZ_TYPE",
  "CZ_FIPS",
  "WFO",
  "INJURIES_INDIRECT",
  "DEATHS_INDIRECT",
  "FLOOD_CAUSE",
  "TOR_LENGTH",
  "TOR_WIDTH",
  "BEGIN_RANGE",
  "BEGIN_AZIMUTH",
  "END_RANGE",
  "END_AZIMUTH",
  "END_LOCATION",
  "END_TIME"
  )
```

```{r}
# Dropping the useless columns from storm

storm_dataset <- storm_dataset[, !(names(storm_dataset) %in% storm_dataset_cols_to_drop)]
#storm_dataset
```

```{r}
# Make sure the date columns are date objects 

storm_dataset <- storm_dataset %>%
  mutate(BEGIN_DATE = mdy(BEGIN_DATE))
#storm_dataset

OBA_dataset <- OBA_dataset %>%
  mutate(verbatimEventDate = mdy(verbatimEventDate))
#OBA_dataset
```

```{r}
# Join Tables

joined_data <- fuzzy_left_join(OBA_dataset, storm_dataset,
  by = c(
    "county" = "COUNTY",                        
    "verbatimEventDate" = "BEGIN_DATE"             
  ),
  match_fun = list(`==`, function(bee_date, storm_date) {
    bee_date >= storm_date & bee_date <= storm_date + 30
    }
  )
)

joined_data
```

```{r}
# Count of each unique storm that occurred before a bee capture instance
joined_data %>%
  count(EVENT_TYPE)
joined_data
```


```{r}
# Add Control column. True if control data, false if a storm instance
joined_data <- joined_data %>%
  mutate(Control = is.na(EVENT_TYPE))
joined_data
```


```{r}
# Remove rows with no species data

joined_data <- joined_data %>%
  filter(specificEpithet != "")
joined_data <- joined_data %>%
  filter(specificEpithet != " ")
```
## Data Analysis - AB Test for Storm vs No Storm

Before we do our AB test to determine whether extreme weather impacts bee diversity, we need to come up with a null hypothesis to either reject or not. Or null hypothesis will be: Bees are equally diverse, according to the shannon diversity index, in areas which have recently been impacted by either extreme heat or freezing temperatures. If we get a p value of less than 0.05, we will reject this null and assume that the weather does impact bees. 

```{r}
# Find Species Richness
S <- length(unique(joined_data$specificEpithet))
S
```


```{r}
# Natural Log of Species Richness 
ln_s <- log(S)
ln_s
```


```{r}
# Calculate Proportion of Each Species

species_counts <- joined_data %>%
  count(specificEpithet, name = 'n')

proportions <- c()

j <- 1

for (i in species_counts$n) {
  proportions[j] = (i / sum(species_counts$n))
  j <- j+1
}
proportions

```


```{r}
# Calculate Overall Shannon Diversity Index
# H = – Summation[P(i) * lnP(i)].

summation <- 0
for (p in proportions) {
  summation <- summation + (p * log(p))
}

shannon_diversity_index <- (summation * -1)

```


```{r}
# Calculate Shannon Diversity Index Another Way

library(vegan)

shannon_diversity_index_2 <- joined_data %>%
  summarize(diversity = diversity(table(specificEpithet)))
shannon_diversity_index_2

```


```{r}
# Calculate Shannon Diversity for Control Group

joined_data
control_table <- joined_data %>%
  filter(Control == TRUE)

S_control <- length(unique(control_table$specificEpithet))

species_counts_control <- control_table %>%
  count(specificEpithet, name = 'n')

proportions_control <- c()

j <- 1

for (i in species_counts_control$n) {
  proportions_control[j] = (i / sum(species_counts_control$n))
  j <- j+1
}
proportions_control

summation_control <- 0
for (p in proportions_control) {
  summation_control <- summation_control + (p * log(p))
}

shannon_diversity_index_control <- (summation_control * -1)

shannon_diversity_index_control
```


```{r}


```


```{r}
# Calculate Shannon Diversity of Treatment Group
joined_data
not_control_table <- joined_data %>%
  filter(Control == FALSE)

S_not_control <- length(unique(not_control_table$specificEpithet))

species_counts_not_control <- not_control_table %>%
  count(specificEpithet, name = 'n')

proportions_not_control <- c()

j <- 1

for (i in species_counts_not_control$n) {
  proportions_not_control[j] = (i / sum(species_counts_not_control$n))
  j <- j+1
}
proportions_not_control

summation_not_control <- 0
for (p in proportions_not_control) {
  summation_not_control <- summation_not_control + (p * log(p))
}

shannon_diversity_index_treatment <- (summation_not_control * -1)

shannon_diversity_index_treatment
```
```{r}
# Create function to compute Shannon Diversity (reproducible)

shannon_diversity <- function(df, col) {
  col <- deparse(substitute(col))

  S <- length(unique(df[[col]]))
  
  species_counts <- df %>%
    count(.data[[col]], name = 'n')
  
  proportions <- c()
  
  j <- 1

  for (i in species_counts$n) {
    proportions[j] = (i / sum(species_counts$n))
    j <- j+1
  }
  
  summation <- 0
  for (p in proportions) {
    summation <- summation + (p * log(p))
  }

  H <- (summation * -1)
  
  return(H)
  
}

shannon_diversity(not_control_table, specificEpithet)

```
So, the Shannon diversity for "storm bees" was 4.3852, and the shannon diversity for bees captured in normal circumstances was 4.646. The diversity is lower for bees captured around storms (Extreme heat or freezing temperatures), but is it significantly so? We will use A/B testing to determine if so.


```{r}
# Set seed for reproducibility, and the number of iterations to shuffle
set.seed(1)   
n_iterations <- 1000

```




```{r}
# Perform AB Testing

shuffle_results <- c()

for (i in 1:n_iterations) {
  shuffled_col <- sample(joined_data$Control)
  shuffled <- joined_data %>%
    mutate(shuffled_control = shuffled_col)
  shuffled_treatment <- shuffled %>%
    filter(shuffled_control == FALSE)
  
  
  shuffle_results[i] <- shannon_diversity(shuffled_treatment, specificEpithet)
}

sort(shuffle_results)


```

```{r}
# Assign P Value (Probability that our observed shannon diversiy for storm bees is at least that small if Null were true)

p_value <- mean(shuffle_results <= shannon_diversity(not_control_table, specificEpithet))
p_value
```
A P value of 0 means that we reject the null hypothesis. This means that the extreme weather events do impact bee diversity, making the bees less diverse in areas that recently experienced either extreme heat or freezing temperatures. 



```{r}
# Histogram of Shuffled Results
ggplot(data.frame(x = shuffle_results), aes(x)) +
  geom_histogram() +
  ggtitle("Distribution of Shuffled AB Test Shannon Diversities") 


```

```{r}
# Histogram of Shuffled Results, plus vertical line showing non-shuffled diversity
ggplot(data.frame(x = shuffle_results), aes(x)) +
  geom_histogram() +
  ggtitle("Distribution of Shuffled AB Test Shannon Diversities + Line for Observed Statistic") + 
  geom_vline(xintercept = shannon_diversity(not_control_table, specificEpithet), color = "red", linetype = "dashed", size = 1)

```
As we can see, our observed statistic (Shannon diversity of storm bees) falls well outside of the limits of our shuffled results, meaning it is statistically significant.

## Data Analysis: Does freezing temperatures impact bee diversity?

Now, we want to do another AB test, but this time just to see if frost/freeze impacts bee diversity (ignoring extreme heat). Ou rnull hypothesis this time will be: Freezing temperatures do not impact bee diversity according to the Shannon diversity index. 

```{r}
# Create new table to analyze freezing temperatures

joined_data_frost <- joined_data %>%
  mutate(frost = coalesce(EVENT_TYPE == "Frost/Freeze", FALSE))
joined_data_frost
```

```{r}
# Calculate Observed Statistic
joined_data_frost_filtered <- joined_data_frost %>%
  filter(frost == TRUE)
shannon_diversity(joined_data_frost_filtered, specificEpithet)
```


```{r}
# Perform AB Test
shuffled_frost_results <- c()
for (i in 1:1000) {
  joined_data_frost <- joined_data_frost %>%
    mutate(frost_shuffled = sample(frost))
  temp <- joined_data_frost %>%
    filter(frost_shuffled == TRUE)
  shuffled_frost_results[i] <- shannon_diversity(temp, specificEpithet)
}

sort(shuffled_frost_results)
```


```{r}
# Compute P Value for freezing temperatures
p_value_frost <- mean(shuffled_frost_results <= shannon_diversity(joined_data_frost_filtered, specificEpithet))
p_value_frost
```

```{r}
# Histogram for frost results
ggplot(data.frame(x = shuffled_frost_results), aes(x)) +
  geom_histogram() +
  ggtitle("Distribution of Shuffled Frost Results + Line for Observed Statistic") + 
  geom_vline(xintercept = shannon_diversity(joined_data_frost_filtered, specificEpithet), color = "red", linetype = "dashed", size = 1)


```
As we see here, the p value is once again 0. This means that we reject the null, and that freezing temperatures does negatively impace the shannon diversity of bees.

## Data Analysis - AB Test for Extreme Heat
For our last AB test, we will look at whether or not extreme heat impacts bee diversity. Our null hpyothesis: Extreme heat events do not impact the diversity of bees. 
```{r}
# Create new table to analyze extreme heat

joined_data_heat <- joined_data %>%
  mutate(heat = coalesce(EVENT_TYPE == "Excessive Heat", FALSE))
joined_data_heat
```

```{r}
# Calculate Observed Statistic
joined_data_heat_filtered <- joined_data_heat %>%
  filter(heat == TRUE)
shannon_diversity(joined_data_heat_filtered, specificEpithet)

```

```{r}
# Perform AB Test
shuffled_heat_results <- c()
for (i in 1:1000) {
  joined_data_heat <- joined_data_heat %>%
    mutate(heat_shuffled = sample(heat))
  temp <- joined_data_heat %>%
    filter(heat_shuffled == TRUE)
  shuffled_heat_results[i] <- shannon_diversity(temp, specificEpithet)
}

sort(shuffled_heat_results)

```

```{r}
p_value_heat <- mean(shuffled_heat_results <= shannon_diversity(joined_data_heat_filtered, specificEpithet))
p_value_heat

```
```{r}
ggplot(data.frame(x = shuffled_heat_results), aes(x)) +
  geom_histogram() +
  ggtitle("Distribution of Shuffled Heat Results + Line for Observed Statistic") + 
  geom_vline(xintercept = shannon_diversity(joined_data_heat_filtered, specificEpithet), color = "red", linetype = "dashed", size = 1)

```
Once again, we reject the Null hypothesis. We also see based on the difference in shannon diversities, that heat actually has more impact on the diversity than the extreme cold!

```{r}
diversities <- data.frame(
  Category = c("Storm"),
  Frost = shannon_diversity(joined_data_frost_filtered, specificEpithet),
  Heat = shannon_diversity(joined_data_heat_filtered, specificEpithet),
  "No Storm" = shannon_diversity_index_control
)

diversities_long <- diversities %>%
  pivot_longer(cols = -Category, names_to = "Type", values_to = "Value")

ggplot(diversities_long, aes(x = Type, y = Value, fill = Type)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  labs(title = "Diversity by Category", x = "Category Type", y = "Shannon Diversity")



```









```{r}
# Find dates of events
Event_Date_Data <- joined_data %>%
  filter(!is.na(EVENT_TYPE)) %>%
  select(BEGIN_DATE, EVENT_TYPE) %>%
  distinct()
```

```{r}
# Create subsets of bees 30-days pre/post events and compute Shannon diversity for each
n_events <- nrow(Event_Date_Data)

pre_div  <- numeric(n_events)
post_div <- numeric(n_events)

for (i in 1:n_events) {
  event_day <- Event_Date_Data$BEGIN_DATE[i]
  
  # Bees 30 days before
  pre_bees <- joined_data %>%
    filter(verbatimEventDate >= event_day - 30,
           verbatimEventDate <  event_day)
  
  # Bees 30 days after
  post_bees <- joined_data %>%
    filter(verbatimEventDate >  event_day,
           verbatimEventDate <= event_day + 30)
  
  # Counts per species
  pre_tab  <- table(pre_bees$GenusSpecies)
  post_tab <- table(post_bees$GenusSpecies)
  
  # Calculate Shannon diversity
  pre_div[i]  <- diversity(pre_tab)
  post_div[i] <- diversity(post_tab)
}
```


```{r}
# Comparison Table
diversity_results <- data.frame(EVENT_DATE = Event_Date_Data$BEGIN_DATE,
                                EVENT_TYPE      = Event_Date_Data$EVENT_TYPE,
                                shannon_pre30   = pre_div,
                                shannon_post30  = post_div,
                                diff_post_minus_pre = post_div - pre_div)

diversity_results
```

```{r}
# Find the means for each event type
event_type_summary <- diversity_results %>%
  group_by(EVENT_TYPE) %>%
  summarise(mean_pre  = mean(shannon_pre30),
            mean_post = mean(shannon_post30),
            mean_difference = mean(diff_post_minus_pre),
            n_events = n())

event_type_summary
```
```{r}
# Bar graph showing differences in pre/post event, grouped by event type

```

Bees were classified as specialists or generalists using genus-level traits derived from established ecological references (Michener 2000; Wilson & Carril 2015; Danforth et al. 2019). Genera with strong evidence of specialization (e.g., Diadasia, Peponapis, Perdita, Chelostoma, Dufourea) were categorized as specialists based on confirmed associations documented by Fowler & Droege (2020) and Praz et al. (2018) and genera (e.g., Bombus, Lasioglossum, Halictus) were categorized as generalists through documentations by Wilson & Carril (2015) and Goulson (2010). Bee identities were additionally cross-referenced with the USGS Native Bee Inventory dataset (Fowler & Droege 2020), which is a widely used North American bee monitoring program. Genera were only assigned to specialist or generalist categories when consistent agreement existed in the literature; ambiguous genera were excluded from assignment.

```{r}
# Create object vectors for Genera dependent on if they are specialists or generalists. Genera that have species that are both are excluded for clarity
# Remove whitespace
joined_data$Genus <- trimws(sub(" .*", "", joined_data$Species))

specialist_genera <- c("Diadasia", "Peponapis", "Hesperapis", "Dufourea", "Chelostoma", "Perdita", "Colletes", "Melissodes")

generalist_genera <- c("Bombus","Halictus","Apis","Lasioglossum","Eucera","Ceratina","Agapostemon",
                       "Xylocopa","Osmia","Anthophora","Megachile","Hylaeus","Calliopsis","Heriades",
                       "Ashmeadiella","Atoposmia","Trachusa","Dianthidium","Tetraloniella",
                       "Pseudoanthidium","Protosmia","Anthidiellum","Hoplitis","Habropoda")
# For clarity
excluded_genera <- c("Andrena","Micralictoides","Stelis","Sphecodes","Melecta","Coelioxys","Epeolus",
                     "Neolarra","Oreopasites","Nomia","Triepeolus","Biastes","Epimelissodes",
                     "Protohalonia","Dioxys","Zacosmia","Holcopasites","Brachymelecta")
```


# Results


# References (Temporary spot. Needs to be changed to BibTex)

Bilton, D. T., McGill, R. A. R., & Parker, D. J. (2019). Extreme events drive range contraction in a cold-adapted species. Global Change Biology, 25(11), 3576–3587.
Franzén, M., & Öckinger, E. (2012). Climate-driven changes in pollinator communities: The return of the dark side of cold winters. Oikos, 121(3), 437–442.
Goulson, D., Nicholls, E., Botías, C., & Rotheray, E. L. (2015). Bee declines driven by combined stress from parasites, pesticides, and lack of flowers. Science, 347(6229), 1255957.
Heinrich, B. (1979). Bumblebee economics. Harvard University Press.
IPCC. (2021). Climate change 2021: The physical science basis. Contribution of Working Group I to the Sixth Assessment Report of the Intergovernmental Panel on Climate Change. Cambridge University Press.
Oregon Climate Change Research Institute (OCCRI). (2023). The Fifth Oregon Climate Assessment. Oregon State University.
Pereboom, J. J., & Biesmeijer, J. C. (2003). Thermal constraints for stingless bee foragers: The importance of body size and coloration. Oecologia, 137(1), 42–50.
Potts, S. G., Imperatriz-Fonseca, V., Ngo, H. T., et al. (2016). The assessment report of the Intergovernmental Science-Policy Platform on Biodiversity and Ecosystem Services (IPBES) on pollinators, pollination, and food production. IPBES Secretariat.
Riley, D. G., Browning, L. E., & Withrow, J. R. (2018). Ice storms, overwintering survival, and pollinator decline: Evidence from temperate ecosystems. Ecology Letters, 21(4), 650–659.
Thornton, P. K., Whitfield, S., & Challinor, A. J. (2023). Climate extremes and their implications for global food systems. Annual Review of Environment and Resources, 48, 203–229.
Woods, H. A., Dillon, M. E., & Pincebourde, S. (2019). The roles of microclimatic diversity and of behavior in mediating the responses of ectotherms to climate change. Journal of Thermal Biology, 84, 80–92.
